Compositional equivalences based on Open pNets\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-12-11 2016-Weak-FH-Bisim/WeakBisim.tex
quelques TODO{commentaires} faits

%Ludo
%\begin{itemize}
%\item[x] rewrite Tr2 and open transition with actions
%\item[x] $v'\to \alpha'$ dans pred sv et enlever l exists
%\item[x] $\beta$ ou $b$
%\item[x] propager Tr2 dans l annexe
%\item[x] add simple equivalence on OT -- fait a minima 
%\item[x] explain the 2 level logic
%\end{itemize}
%=== 18 Avril
\medskip

Eric
\begin{itemize}
\item[X] nettoyage (1/2) -> 5 juil
\item nettoyage exemple et algo de calcul
\item example -> fin aout -> qq transitions dans l abp
%\item re-ecrire transitions page 14 comment les presenter?
%\item arbre de deduction
%\item exemple avec actions
\end{itemize}

Rabea
\begin{itemize}
\item preuve transitivite -> 20 sept
\item proofs of new theorems -> oct?
%\item [x] uniformiser p et P pour les pnets 
%\item [x] relectures
\end{itemize}
=== en parallele

@@

We denote $y\gets e$ a substitution. The application of the substitution is denoted
$\subst{y\gets e}$, the operation replaces in a term all occurrences 
of the variable $y$ by the expression $e$ (note that the absence of binders makes this operation 
trivial). 
We use sets of equations of the form $x_i=e_i$, ranged over by $\Post$; $\subst{\Post}$ is the substitution that applies all the substitutions defined by $\Post$ in a parallel manner. $\otimes$ is the composition operator on substitutions: $(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$

@@

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on variables (potentially with only symbolic values). Formaly, each state of an OA has a set of \emph{state variables}, and all state variable sets are disjoint.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$ and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
        We call \emph{source variables} of an OT the union of all variables
        in the different terms
        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the source variables. \Post\ is a set of 
	equations that hold \emph{after the open transition}, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $x_k$ are state variables of $s'$, and $e_k$ are expressions
        over the source variables. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

%\TODO{Eric: rewrite example with new Protocolkeep only OT1, and adapt comments}
\begin{example}\emph{An open-transition.}
  \label{OT:ABP}
At the upper level, the $SimpleSystem$ pNet of Figure \ref{SimpleProt:Impl} has 2 holes and $SimpleProtocol$ as
a subnet, itself containing 3 pLTSs. One of its possible open transitions
(synchronizing the hole $P$
with the $Sender$ within the \emph{SimpleProtocol}) is:

 \smallskip\noindent
 \[  OT_1  = \openrule{
      \{P\mapsto p\_send(m)\} \qquad [m=m'] \qquad \{s1\_m:=m\}
                      }
    {\ostate{s_0,a_0,r_0} \OTarrow{\underline{in(m')}} \ostate{s_1,a_0,r_0}}
    \]

    \smallskip
    The global states here are triples, build as the product of states of the 3 pLTS (remember the Holes have no state). The assignment
    $Post$ uses a variable from the action of hole $P$ to set the
    value of the sender state variable $s1\_m$.
    \TODO{need more comments ?}

@@

Figure \ref{SimpleProtCounter:ImplOA} shows the full open automaton of the implementation $SimpleSystem$. Here, in the open automaton drawing, we use a user-friendly notation for the OTs, in the form ``$\{<actions of holes\} [<guard>] <action> \{assignments\}$''.
\TODO{comment on difference from Exemple 1 page 15; and predicate simplification...}
\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{ATG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

%% $  OT_1  = \openrule{
%%   (P\mapsto p\_send(m:sva\_SV6:1:1)) \\
%%   [( msg =m: sva\_SV6 :1:1) \wedge (( in( msg ))=( in(m: sva_SV6 :1:1)))]\\
%%   \{ s1\_msg := msg , s1\_c := 0\}
%%                       }
%% {\ostate{s_0,m_0,r_0} \OTarrow{\underline{in(m:sva\_SV6:1:1)}} \ostate{s_1,m_0,r_0}}
%% $
    \TODO{Comment...}

@@

  WOT2 here is the transition delivering the result of the algorithm to hole $Q$. Remark that it includes a loop increasing the counter occurs before the action of $Q$, and this action is using the value of the counter. Let us go into the details, using rules WT1/2/3. For simplification we ignore the pure $\tau$-loops before and after the central \texttt{q-recv(ec)} action, they have no effect on the construction. First consider application of Rule WT3 on the first 3 transitions. Hole actions and predicates are empty, we just concentrate on the Post operations. Rule WT2 transforms each 3 OTs into WOTs. In the drawing we use \texttt{ec++} to abbreviate the Post: $\subst{ec\gets ec+1}$. Then WT3 produces the weak OT:
  $\openrule
         {\set{\tau}, True,
			\subst{\texttt{ec++}\shortotimes\texttt{ec++}\shortotimes\emptyset} }
         {s \OTWeakarrow {\tau\subst{\texttt{ec++}}} s'}$
         that you can rewrite as:      
  $\openrule
         {\emptyset, True,
			\subst{ec\gets ec+2} }
         {s \OTWeakarrow \tau s'}$

@@

\subsection{Running example}
- If we already have shown the saturated relation, we only need to argue on bisimulation here.

@@

The ``saturation'' step in the construction of the weak relation will
pick all $\tau^*$ and $\tau^*;a;\tau^*$ sequences from the OA, and build a weak
open trnsition from each sequence. Here is one example of a short sequence,
starting with the open transition $OT_2$ of page \pageref{OT:ABP},
then transmitting the message to the receiver (if the sender and
receiver control bits are equal), and to the environment. The two
firsts open transitionTs are internal, the last one is visible, so it constitute a
$\tau;\tau;a$ weak transition.

\TODO{Eric: rewrite completely}
\noindent
     $  \openrule{
      \emptyset; Pred=\True; Post=\{m_1\_m:=s_1\_m, m_1\_b:=s_1\_sb, s_2\_sb:=s_1\_sb\}
                      }
    {\ostate{s_1,m_0,m_0,r_0} \OTarrow{\underline{send(s_1\_m, s_1\_b)}} \ostate{s_2,m_1,m_0,r_0}}
    $

@@

    \ERIC{Apres simplification:}
\medskip\noindent
     $  \openrule{
      \emptyset; Pred=[r_0\_rb=m_1\_b];
      \\ Post=\{r_1\_rb:=r_0\_rb; r_1\_m:=m_1\_m; r_1\_b:=m_1\_b      \}
                      }
    {\ostate{s_2,m_1,m_0,r_0} \OTarrow{\underline{recv(m_1\_m,m_1\_b)}} \ostate{s_2,m_0,m_0,r_1}}
    $
    \\

>>>>

@@

We denote $y\gets e$ a substitution. The application of the substitution is denoted
$\subst{y\gets e}$, the operation replaces in a term all occurrences 
of the variable $y$ by the expression $e$ (note that the absence of binders makes this operation 
trivial). 
We use \ERIC{sets of assignments of the form $x_i:=e_i$}, ranged over by $\Post$; $\subst{\Post}$ is the substitution that applies all the substitutions defined by $\Post$ in a parallel manner. $\otimes$ is the composition operator on substitutions: $(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$

@@

\subsection{Open Automata}
 Open automata (OA) are not composition structures but they are made of transitions that are dependent of the actions of the holes, and they can reason on variables (potentially with only symbolic values). Formaly, each state of an OA has a set of \emph{state variables}, and all state variable sets are disjoint.
%\TODO{adopt a uniform notation for open transitions, almost each instance has a 
%different 
%notation! I suggest p,l,pr,po using \{\} for p,l,po as they are sets}
\begin{definition}[Open transitions]\label{def:OT}
	\label{def:OpenTransitions}
	An \emph{open transition} (OT) over a
	set $J$ of holes with sorts $\Sort_j^{j\in J}$ and a set of states $\mathcal{S}$ is 
	a structure of the form:	
	\begin{mathpar}
	\openrule
	{	\beta_j^{j\in J'}, \Pred, \Post}
	{s \OTarrow {\alpha}s'}
	\end{mathpar}
	Where $J'\subseteq J$, $s, s'\in\mathcal{S}$ and $\beta_j$
        is a transition of the hole $j$, with $\beta_j\in\Sort_j$. $\alpha$ is an action 
        label denoting the resulting action of this open transition.
        We call \emph{source variables} of an OT the union of all variables
        in the different terms
        $\beta_j$ and $\alpha$, and the state variables of $s$.
        \Pred\ is a predicate over the source variables. \Post\ is a set of 
	\ERIC{assignments that are effective \emph{after the open transition}}, they are
        represented as a substitution of the form $({x_k\gets e_k})^{k\in K}$ 
	where $x_k$ are state variables of $s'$, and $e_k$ are expressions
        over the source variables. Open transitions are identified
        modulo logical equivalence on their predicate. 
\end{definition}

@@

%\TODO{Eric: rewrite example with new Protocolkeep only OT1, and adapt comments}
\begin{example}\emph{An open-transition.}
  \label{OT:SimpleProt}
At the upper level, the $SimpleSystem$ pNet of Figure \ref{SimpleProt:Impl} has 2 holes and $SimpleProtocol$ as
a subnet, itself containing 3 pLTSs. One of its possible open transitions
(synchronizing the hole $P$
with the $Sender$ within the \emph{SimpleProtocol}) is:

 \smallskip\noindent
 \[  OT_1  = \openrule{
      \{P\mapsto p\_send(m)\} \qquad [m=m'] \qquad \{s1\_m\gets m\}
                      }
    {\ostate{s_0,a_0,r_0} \OTarrow{\underline{in(m')}} \ostate{s_1,a_0,r_0}}
    \]

    \smallskip
    The global states here are triples, build as the product of states of the 3 pLTS (remember the Holes have no state). The assignment
    $Post$ uses the variable $m$ from the action of hole $P$ to set the
    value of the sender state variable named $s1\_m$.

@@

Figure \ref{SimpleProtCounter:ImplOA} shows the full open automaton of the implementation $SimpleSystem$. Here, in the open automaton drawing, we use a user-friendly notation for the OTs, in the form ``$\{actions of holes\} [guard] action \{assigns\}$'', rather than the full OT form with the red-dotted rule. Note that the resulting action of the OTs appear here betwen the guard and the set of assignments.
Note that we have also simplified the form of the variable dependencies in the OTs, again for readability, replacing e.g. in OT1 $\{P\mapsto p\_send(m)\} \qquad [m=m']$ and $\alpha= \underline{in(m')}$ by the shorter \texttt{\{P->p\_send(m)\} in(m)}.

    \begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{ATG/SPImplOpen}}
  \caption{Open Automaton of the Simple Protocol Implementation}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

%% $  OT_1  = \openrule{
%%   (P\mapsto p\_send(m:sva\_SV6:1:1)) \\
%%   [( msg =m: sva\_SV6 :1:1) \wedge (( in( msg ))=( in(m: sva_SV6 :1:1)))]\\
%%   \{ s1\_msg := msg , s1\_c := 0\}
%%                       }
%% {\ostate{s_0,m_0,r_0} \OTarrow{\underline{in(m:sva\_SV6:1:1)}} \ostate{s_1,m_0,r_0}}
%% $
    \ERIC{Also we have short labels for states, representing $\ostate{s_0,m_0,r_0}$ by \texttt{000}. The resulting behavior is quite simple: In every state, holes $P$ and $Q$ can do any action independent of the protocol (we only have labelled these on state \texttt{000}). Then we have a main loop including receiving a message from $P$ and transmitting the same message to $Q$, with some intermediate $\tau$ actions from the internal communications between the protocol processes. In most of the transitions, you can observe that data is propagated between the successive state variables (holding the message, and the error counter value). On th eright of the figure, you have a loop of $\tau$ transitions showing the handling of errors and corresponding increments of the error counter.}
    \TODO{j'ai comment\'e Fig 3, vous pouvez relire ?}

@@

  WOT2 here is the transition delivering the result of the algorithm to hole $Q$. Remark that it includes a loop increasing the counter, occuring before the action of $Q$, and this action is using the value of the counter. Let us go into the details, using rules WT1/2/3. For simplification we ignore the pure $\tau$-loops before and after the central \texttt{q-recv(ec)} action, they have no effect on the construction. First consider application of Rule WT3 on the first 3 transitions. Hole actions and predicates are empty, we just concentrate on the Post operations. Rule WT2 transforms each 3 OTs into WOTs. In the drawing we use \texttt{ec++} to abbreviate the Post: $\subst{ec\gets ec+1}$. Then WT3 produces the weak OT:
  $\openrule
         {\set{\tau}, True,
			\subst{\texttt{ec++}\shortotimes\texttt{ec++}\shortotimes\emptyset} }
         {s \OTWeakarrow {\tau\subst{\texttt{ec++}}} s'}$
         that you can rewrite as:      
  $\openrule
         {\emptyset, True,
			\subst{ec\gets ec+2} }
         {s \OTWeakarrow \tau s'}$

@@

\subsection{Running example}
- In section \ref{} we have shown the full satured weak automaton for both the specification and the implementation of the simple protocol.

@@

  \TODO{Eric: this is still ABP ! rewrite completely, baesd on saturation example 2 pages before.}

The ``saturation'' step in the construction of the weak relation will
pick all $\tau^*$ and $\tau^*;a;\tau^*$ sequences from the OA, and build a weak
open transition from each sequence. Here is one example of a short sequence,
starting with the open transition $OT_1$ of page \pageref{OT:SimpleProt},
then transmitting the message to the receiver, and to the environment. The two
firsts open transitionTs are internal, the last one is visible, so it constitute a
$\tau;\tau;a$ weak transition.

noindent
     $  \openrule{
      \emptyset; Pred=\True; Post=\{m_1\_m:=s_1\_m, m_1\_b:=s_1\_sb, s_2\_sb:=s_1\_sb\}
                      }
    {\ostate{s_1,m_0,m_0,r_0} \OTarrow{\underline{send(s_1\_m, s_1\_b)}} \ostate{s_2,m_1,m_0,r_0}}
    $

@@

    \ERIC{Apres simplification:}

\medskip\noindent
     $  \openrule{
      \emptyset; Pred=[r_0\_rb=m_1\_b];
      \\ Post=\{r_1\_rb:=r_0\_rb; r_1\_m:=m_1\_m; r_1\_b:=m_1\_b      \}
                      }
    {\ostate{s_2,m_1,m_0,r_0} \OTarrow{\underline{recv(m_1\_m,m_1\_b)}} \ostate{s_2,m_0,m_0,r_1}}
    $
    \\
